---
title: "NYC transit EDA"
author: "Ravi Brenner"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(jsonlite)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

The idea here is to explore changes in transit and transportation following the implementation of congestion pricing. Essentially, I want to know if people are choosing alternative modes of transportation in place of driving.

My initial thought is to mainly use maps, line/bar graphs, and maybe some animated graphs to show what is happening.

A further project/idea is to incorporate weather (esp into bike usage) and try to predict (i.e. using regression or ML models) transit use given different conditions.

Also, I would ideally want this to end up as a dashboard that could be updated in real time (absolutely inspired by <https://www.congestion-pricing-tracker.com/>).

Datasets I will be using:

-   [Bicycle Counts](https://data.cityofnewyork.us/resource/uczf-rk3c)

-   [MTA Subway Hourly Ridership: Beginning July 2020](https://data.ny.gov/resource/wujg-7c2s)

-   [MTA Bus Hourly Ridership: Beginning September 2020](https://data.ny.gov/resource/kv7t-n8in)

-   [MTA Bridges & Tunnels Hourly Traffic Rates: Beginning 2010](https://data.ny.gov/resource/qzve-kjga)

I also anticipate needing some additional datasets for mapping the data, and for obtaining weather data. Another future direction could be to incorporate regional rail data as well.

Overall, one challenge I forsee is handling all the data. Many of these datasets have millions of rows, which could be hard to pull via the API, and may not be easily stored in memory as R does (maybe a python solution will be better? We will see).

## Bicycle data

Pull a 2 year window of data to limit rows
```{r}
max_date <- fromJSON("https://data.cityofnewyork.us/resource/uczf-rk3c.json?$select=max(date)") |>
  pull() |>
  str_replace("T"," ") |>
  as.POSIXct()

# subtract 2 years and reformat
min_date <- as.character(max_date - dyears(2)) |>
  str_replace(" ","T") |>
  str_c(".000")

bike_rows <- fromJSON(str_c("https://data.cityofnewyork.us/resource/uczf-rk3c.json?$select=count(id)&$where=date>='",min_date,"'")) |>
  pull()

bike_df <- fromJSON(str_c("https://data.cityofnewyork.us/resource/uczf-rk3c.json?$where=date>='",min_date,"'&$limit=",bike_rows)) |>
  as_tibble()
```

Bicycle counters
```{r}
counters_df <- fromJSON("https://data.cityofnewyork.us/resource/smn3-rzf9.json") |>
  as_tibble()
```

merge
```{r}
bike_df <- bike_df |>
  inner_join(counters_df,by = "id")  
```

transform columns
```{r}
bike_df <- bike_df |>  
  mutate(counts = as.numeric(counts),
         date = str_replace(date,"T"," "),
         date = str_remove(date,".000"),
         date = as.POSIXct(date),
         year = year(date),
         month = month(date),
         day = day(date),
         wday = wday(date),
         hour = hour(date))
```

Track by location
```{r}
bike_df |>
  summarize(.by = c(name, year, month, day, hour),
            counts = sum(counts)) |>
  mutate(my_date = ymd_h(paste(2024, month, day, hour, sep = " "))) |>
  filter(month == 1) |>
  filter(name == "Amsterdam Ave at 86th St.") |>
  ggplot(aes(x = my_date, y = counts, color = factor(year))) + 
  geom_line()
```



## Subway data
```{r}
max_dates <- fromJSON(str_c("https://data.ny.gov/resource/wujg-7c2s.json?$query=",URLencode("SELECT max(transit_timestamp)"))) |>
  pull() |>
  str_replace("T"," ") |>
  as.POSIXct()

# subtract 2 years and reformat
min_date <- as.character(max_date - dyears(2)) |>
  str_replace(" ","T") |>
  str_c(".000")


query = URLencode(str_c("SELECT transit_timestamp, station_complex_id, sum(ridership) GROUP BY transit_timestamp, station_complex_id"))

subway_df <- fromJSON(str_c("https://data.ny.gov/resource/wujg-7c2s.json?$query=",query)) |>
  as_tibble()


```


## Bus data

## Driving data

